### 2020-01-30 paper/sex_selection_spacing-ver3.tex
Two tables for descriptive stats
Editing education and conceptual framework sections

India has changed significantly over the last four decades:
the economy has grown substantially,
education levels have increased for both males and females,
and the total fertility rate has fallen to the point where, at 2.2, it is just
above replacement 
\citep{Bosworth2008,Guilmoto2013,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

Over the same period, India has, however, also experienced a rapidly expanding 
access to prenatal sex determination.
Combined with continued strong son preference, the result has been a dramatic 
increase in the males-to-females ratio at birth
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,
Bongaarts2013,Portner2015b,Jayachandran2017}.

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions has the potential to substantially increase birth spacing.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Add to this increase that the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries as they developed.

@@

First, tempo effects, here arising from changes in birth spacing, bias 
the total fertility rate (TFR), in the sense that the TFR becomes a more biased measure 
of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
If birth spacing increased substantially over time, India's fertility may, in 
fact, be higher than what is generally accepted based on the TFR.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Maitra2008,Makepeace2008,
Jayachandran2011,Kumar2016,Jayachandran2017a,Ghosh2018}.
Longer spacing between births then potentially lowers mortality, although there may
be a counteracting effect of multiple abortions.

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons, rather than an indicator of 
lower son preference, may instead arise because couples with a strong preference 
for sons use sex selection, possibly to the point where we observe a reversal of 
the traditional spacing pattern.
Despite a large literature on son preference and birth spacing, there is, however,
no prior research that has examined how sex selection affects spacing.

@@

Second, how has the distribution of birth spacing changed?
Although examiniong average birth intervals helps us understand the overall changes in 
spacing behavior, it tells us little about how these changes came about.
For example, the motivations that lead to an increase in average spacing are different if 
the increase comes mainly from parents trying to avoid very short birth intervals or from 
continued use of sex selection to ensure a son.

[TK this is unclear]
The traditional pattern of longer average spacing with sons appears to arise from 
an increase in the number of very long intervals, rather than from fewer very short 
birth intervals. 
Across education groups, there appears to be little difference in the likelihood of 
having a very short birth interval across sex compositions, especially for lower parities.

@@

\section{Birth Spacing: Mechanisms and Findings}

This section provides a conceptual framework for understanding what determines  
birth spacing.
I first discuss the main determinants of spacing examined in the prior research, and 
then how son preference and sex selection fits into the spacing decisions.
Since sex selection and education are closely linked, I pay special attention to how 
education affects spacing.

In the simplest possible dynamic model of fertility, the spacing decision is trivial.
Assume that parents derive benefits from the presence of children, not their births, 
that spacing of children does impact parents' utility, 
and that the parents fully control when to have children. 
The optimal decision then is to have the children as early as possible 
\citep{Newman1988}.
The simplest model does not fit the observed spacing patterns, which suggests 
that one or more of the assumptions are wrong.

First, parents do not entirely control the timing of births, but rather decide on 
the levels of sexual activity and contraceptive effort, which, in turn, affect the 
likelihood of conceiving, while taking into account fecundity and their knowledge 
of and access to contraceptive technology.

Fecundity determinants, such as age and health of the mother, are, however, unlikely 
to be significant factors in how spacing changed over time or across education levels 
in India for the period covered.
Outside of direct famine conditions, women's health and nutritional status do not 
appear to affect spacing substantially \citep{Huffman1987,John1987,lindstrom99}.
The Maharashtra drought occurred in 1970--1973 and would, therefore, only impact the first
year or two of the data.
Furthermore, I only examine spacing up to parity fourth, and the age of the mother is,
therefore, unlikely to be a concern when it comes to fecundity, even though 
better-educated women are, on average, older when they marry.

Despite a general belief that better access to and knowledge of contraception 
should lengthen average spacing, the empirical relationship between contraception 
use and birth spacing is ambiguous 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
This lack of evidence is consistent with the ambiguous theoretical effect of introducing 
modern contraceptives on birth spacing.
On the one hand, in the absence of high-efficacy contraceptive technologies, parents may 
try to avoid overshooting their desired number of children by spacing their children 
further apart than they would otherwise do.
Increased reliability of the access to and effectiveness of contraceptives then lead to 
shorter birth intervals because parents are more confident they can effectively
stop childbearing when they want \citep{Keyfitz1971,Heckman1976}.
On the other hand, through simple random variation, when contraceptives are ineffective, 
some birth intervals will be shorter than ideal, in the way discussed below.
With increased reliability of contraceptive efforts, parents can better avoid too short 
birth spacing, and average spacing should go up.

To further complicate the relationship between contraception and spacing, the ability 
to successfully use ``low efficacy'' contraceptive methods, such as the rhythm method, 
is increasing in education, while there is no difference across education levels 
when it comes to modern contraceptives such as pills or injectables \citep{Rosenzweig1989}.
Additionally, even in the absence of modern contraceptives, families appear able to 
time or at least postpone births when they have substantial enough incentive 
\citep{Jayachandran2011,Alam2018}.

Second, parents may care about the spacing between birth because of the 
potential effects of spacing on the children.%
\footnote{
In principle, the mother could also be affected, but the estimated impact on 
mothers' anthropometric status is mixed, and there appears to be no 
effect on mortality 
\citep{Ronsmans1998,Menken2003,Dewey2007,Conde-Agudelo2012}.
}
One theory is that child ``quality'' increases with spacing 
since parental attention per child increases
\citep{Zajonc1975,Zajonc1976,Razin1980}.
In this case, birth spacing should increase in family income, while the 
effect of maternal education is ambiguous since higher education, on the 
one hand, makes the mother's time spent out of the labor market more 
costly and on the other hand increases how productive the mother is 
in childrearing.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

For one crucial aspect of child quality---health and mortality---shorter spacing 
does lead to worse outcomes, especially for very short intervals of 24 months 
or less, although this effect decreases with maternal education
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
The underlying mortality risk could also affect birth spacing, but
there is little empirical evidence for this 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
If parents realize short birth spacing increases their children's mortality risk, 
they likely prefer to space births further apart.
However, this effect should diminish with education level since women with 
secondary or tertiary education experience relatively low mortality risk for 
their children, even with very short spacing is associated.

Third, parents incur both direct and time costs when they have children, 
and economic considerations can, therefore, affect birth spacing 
\citep{Hotz1997,schultz97}.
I divide the economic factors into the effects of economies of scale and
the expected income trajectory of the husband and wife.

The expected effect of economies of scale in childrearing, which, for example, 
can arise because taking care of two young children does not take twice as much 
time as taking care of one child, is to shorten intervals between births
\citep{Vijverberg1982}.
To the extent that children require the mother to do less market work, parents 
can lower the cost of children by reducing birth spacing.
In this case, better-educated women will both have lower fertility and shorter
spacing than less-educated women because of their higher opportunity cost of time 
\citep{Ross1974,Newman1984}.
However, if the mother mainly works at home or as household income increases, there 
is less incentive to space children for this reason.

In the absence of perfect capital markets, theory predicts that the steeper 
the parents' lifetime income profile, the longer the spacing between births, 
which suggests that with economic development we should see a general increase 
in spacing across education groups and that this increase in spacing should be 
stronger with increasing education levels
\citep{Heckman1976,wolpin84,Newman1988}.
The idea is that although parents would prefer to have their children as early 
as possible to enjoy them for longer, there is also a direct cost to children.
Income is low early on, which means that the marginal utility of consumption is 
higher, making parents less willing to give up consumption to have a child, 
compare to later when they know that their income will be higher 
\citep{Newman1984,Happel1984}.

Finally, the discussion has so far assumed that children are identical---except 
for any differences that arise from spacing---but there is substantial evidence 
that the sex composition of previous children affects birth spacing, with the
absence of boys significantly decreasing the duration to next birth
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.
There is also evidence of differential stopping behavior, with an additional child 
more likely the fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.

If short spacing lowers the survival probability of the next child, as discussed above,
it is not, however, apparent that we should expect shorter spacing in the absence of boys.
There are two potential reasons why parents might still try to conceive sooner 
in the absence of boys.%
\footnote{
The significant differences in spacing by sex composition indicate that
parents control the timing of birth, at least to some extent.
In Matlab, the effect of son preference on birth spacing is stronger in 
areas with better access to family planning \citep{Rahman1993}.
}
One possibility is that parents do not realize that short spacing is potentially
detrimental to the next child's health, resulting in high mortality for both boys 
and girls born after short spacing across all households.

@@

In summary, birth spacing is affected by female education through 
multiple pathways, with the main one of interest here the use of 
sex selection. 
But, to fully understand how sex selection affects spacing, we also 
need to understand the other changes that affect birth spacing over time.
The most likely factors are the changes in household income and 
wealth and the improvements in health as India has developed.
I, therefore, examine how female education and labor force 
participation has changed over time in India.

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
A decline in female labor force participation at the beginning of development is
consistent with the hypothesis of a U-shaped labor force participation as a country
develops \citep{Goldin1994}.
India's female labor force participation is, however, lower than most other
countries and more in line with countries in the Middle East and North Africa, and
does not yet show any signs of increasing
\citep{Klasen2015,Chatterjee2018}.

Theory predicts that increasing women's wages leads to higher labor force participation, 
while both a higher male income and stronger social restrictions on working, possibly in 
specific industries or jobs, reduce female labor supply \citep{Goldin1994}.
India's economy has grown substantially since the mid-seventies, with
an average annual growth rate of about 5.5 percent from 1978 to 2004 \citet{Bosworth2008}.
As a result, real wages for both men and women have almost doubled between 1987 and 2011 
\citep{Klasen2015}.
Despite the increases in wages and the higher female education, the mean male wage is
still close to 70 percent higher than the mean female wage \citep{Bhargava2018}.
Furthermore, the development in women's wages is not uniform across education 
levels \citep{Klasen2015,Bhargava2018}.
The real wages for women with middle school education or below have converged. 
For women with secondary education, the change over time depends on the sample used,
with some showing a substantial decline and others a modest increase.
Women with more than secondary education experienced an increased in their real 
wages, especially in cities.

After controlling for household characteristics and husband income, there is
still a U-shaped relationship between female education and labor supply with the
lowest labor force participation for women with secondary education \citep{Chatterjee2018}.
There are two possible explanations for this effect, both directly related to the strong
pro-male bias in India.
First, it becomes less socially acceptable for women to work in manual labor as their
education increases, and women are still mostly excluded from white-collar
employment despite robust economic growth \citep{Klasen2015,Chatterjee2018}.
Second, with increasing education, women's productivity at home also increases, especially
in the production of offspring human capital.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing male income, demand for better-educated women can 
increase, even if they do not participate in the labor market \citep{Behrman1999}.

Figure \ref{fig:work_by_survey} shows the percent of married women who are 
currently working at the time of the survey by age group and education level.
No other labor force participation question is consistently available across all 
four surveys. 
Because the question refers to currently working, the percentages will be lower 
than what other studies have found. 
Furthermore, since the question is asked only of married women, the overall female labor 
force participation might have developed differently, especially since many young, 
unmarried women have begun working in, for example, the business process 
outsourcing industry, which, in turn, has increased girls' schooling investments
\citep{Jensen2012}.

The U-shaped relationship between education and working holds in most cases with 
the highest percent working for women with either no education or with twelve or 
more years and the lowest for women with eight to eleven years of education.
Women are more likely to report working if they live in rural than urban areas 
and the older they are.
Comparing 1992 and 2015, the percent of women who currently work has either 
remained roughly the same or decreased.
The cases were the numbers were higher in 1999 and 2006 than in 1992 or 2015 
are mostly based on small samples, such as rural women with twelve or more 
years of education in rural areas.

Almost all of the urban women who work receive either cash or a combination of 
cash and in-kind for their work (see Appendix Figure \ref{fig:work_cash_by_survey}). 
Rural women have become more likely to receive cash for their labor over time, 
and the better educated, the more likely they are to receive cash.

Despite this, women across all groups have become substantially more likely to 
work for a family member rather than for themselves or somebody else
(see Appendix Figure \ref{fig:work_family_by_survey}).
Hence, the lower likelihood of working stems mostly from women retracting from 
the general labor market and self-employment.

@@

Women with different education levels have different hazard profiles, although the size 
and direction of the effect vary across areas and time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.
Furthermore, the use of sex selection increases with education, either because of the 
associated lower fertility or because higher income enables them to access and
afford prenatal sex determination
\citep{das_gupta97,retherford03b,jha06,Guilmoto2009a,Bongaarts2013,Portner2015b,Jayachandran2017}.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining son preference \citep{bhat03,pande07}.
}
I divide women into three groups based on education attainment: no
education, one to seven years of education, and eight and more years of education.
The models are estimated separately for each education level to reduce the potential 
problem of including other variables as proportional.

The sex composition of previous children affects both the timing of births and the use of 
sex-selective abortions 
\citep{retherford03b,jha06,Bhalotra2008,abrevaya09,Portner2015b,Kumar2016,Soest2018}.
I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
Finally, urban women have both lower fertility and, presumably, better access to
sex selection, both of which lead to higher use of sex-selective abortions than in 
rural areas \citep{retherford03b,jha06,Portner2015b}.
Area of residence is a dummy variable for the household living in an urban area.

@@

Appendix Table \ref{tab:des_stat1} presents descriptive statistics for
the spells by education level and period.
The level of censoring increases with parity and time, as we should expect with later 
childbearing, falling fertility, and the hypothesized increases in birth intervals from 
sex-selective abortions.

>>>>

India has changed significantly over the last four decades:
the economy has grown substantially,
education levels have increased for both males and females,
and the total fertility rate has fallen to 2.2, or just above replacement 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

India has, however, also experienced a rapidly expanding 
access to prenatal sex determination.
Combined with continued strong son preference, the result has been a dramatic 
increase in the males-to-females ratio at birth
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,Guilmoto2012,
Bongaarts2013,Portner2015b,Jayachandran2017}.

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions has the potential to substantially increase birth spacing.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Furthermore, the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries as they developed.

@@

First, tempo effects, such as those arising from changes in birth spacing, bias 
the total fertility rate (TFR), in the sense that the TFR becomes a more biased measure 
of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
If birth spacing increased substantially over time, India's fertility may, in 
fact, be higher than generally accepted based on the TFR.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhargava2003,Bhalotra2008,Maitra2008,
Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions.
There has so far been no attempt to examine which mechanism dominates.

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons may instead arise because couples 
with a strong preference for sons use sex selection, possibly to the point where we 
observe a reversal of the traditional spacing pattern.
Despite a large literature on son preference and birth spacing, there is, however,
no prior research that has examined how sex selection affects spacing.

@@

Second, how has the distribution of birth spacing changed?
Average birth intervals help us understand the overall changes in 
spacing behavior, but tells us little about how these changes came about.
Examining how the distribution has change allow us to separate out increases in birth
spacing that come from women with strong son preference who still try to conceive early, 
but are also the most likely to end up with very long spells because of multiple 
abortions versus women who have longer average spacing because they try to avoid very
short birth intervals.

[TK this is unclear - complete rewrite]
The traditional pattern of longer average spacing with sons appears to arise from 
an increase in the number of very long intervals, rather than from fewer very short 
birth intervals. 
Across education groups, there appears to be little difference in the likelihood of 
having a very short birth interval across sex compositions, especially for lower parities.

@@

To set the stage for the subsequent analyses, I first describe how female education
and labor force participation has changed.
Both are potentially important factors in the birth spacing decision, as I discuss
in detail in the subsequent section.

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
A decline in female labor force participation at the beginning of development is
consistent with the hypothesis of a U-shaped labor force participation as a country
develops \citep{Goldin1994}.
India's female labor force participation is, however, lower than most other
countries and more in line with countries in the Middle East and North Africa, and
does not yet show any signs of increasing
\citep{Klasen2015,Chatterjee2018}.

Figure \ref{fig:work_by_survey} shows the percent of married women who are 
currently working at the time of the survey by age group and education level.
No other labor force participation question is consistently available across all 
four surveys. 
Because the question refers to currently working, the percentages are lower in previous
studies.

There is a U-shaped relationship between education and working in most cases with 
the highest percent working for women with either no education or with twelve or 
more years and the lowest for women with eight to eleven years of education.
Women are more likely to report working if they live in rural than urban areas 
and the older they are.
Comparing 1992 and 2015, the percent of women who currently work has either 
remained roughly the same or decreased.
The cases were the numbers were higher in 1999 and 2006 than in 1992 or 2015 
are mostly based on small samples, such as rural women with twelve or more 
years of education in rural areas.

\section{Birth Spacing: Mechanisms and Prior Findings}

To understand the potential implications of the substantial changes in income, female 
education, and female labor force participation that India has experience on birth
spacing, this section discuss relevant theoretical frameworks and prior findings 
on birth spacing.

The standard economic argument for why increasing female education has predominantly
been associated with shorter spacing is that parents incur time costs when they have 
children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shorten birth spacing.
With the shorter birth spacing, parents can take advantage of economies of scale in 
childrearing, because taking care of two young children does not take twice as much 
time as taking care of one child \citep{Vijverberg1982}.
In this case, better-educated women will both have lower fertility and shorter
spacing than less-educated women because of their higher opportunity cost of time 
\citep{Ross1974,Newman1984}.

If the mother is unlikely to work in market work, as has become more common over time in 
India, there is little incentive to children more closely together for economic reasons.

% [TK not clear that these two paragraphs add anything to this section]
% The question is then why women appear to be leaving the labor force.
% With the strong economic growth in India, real wages for both men and women have almost 
% doubled between 1987 and 2011 \citep{Klasen2015}.
% Despite the increases in wages and the higher female education, the mean male wage is
% still close to 70 percent higher than the mean female wage and women’s labor supply 
% appear to be more negatively elastic to their husbands' wages than it is positively 
% elastic to own wages, suggesting that the reduction in female labor supply is driven by 
% an income effect \citep{Bhargava2018}.%
% \footnote{
% Furthermore, the development in women's wages is not uniform across education 
% levels \citep{Klasen2015,Bhargava2018}.
% The real wages for women with middle school education or below have converged. 
% For women with secondary education, the change over time depends on the sample used,
% with some showing a substantial decline and others a modest increase.
% Women with more than secondary education experienced an increased in their real 
% wages, especially in cities.
% }

% However, even after controlling for household characteristics and husband income, there 
% still is the U-shaped relationship between female education and labor supply shown
% above with the lowest labor force participation for women with secondary 
% education \citep{Chatterjee2018}.
% There are two possible explanations for this effect, both directly related to the strong
% pro-male bias in India.
% First, it becomes less socially acceptable for women to work in manual labor as their
% education increases, and women are still mostly excluded from white-collar
% employment despite robust economic growth \citep{Klasen2015,Chatterjee2018}.
% Second, with increasing education, women's productivity at home also increases, 
% especially in the production of offspring human capital.
% With relatively more boys born because of increased access to sex-selective 
% abortions and an increasing male income, demand for better-educated women can 
% increase, even if they do not participate in the labor market \citep{Behrman1999}.

With increasing income, parents may also prefer higher child ``quality'' and that would
also predict longer birth intervals since parental attention per child increases
\citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
Nevertheless, for one crucial aspect of child quality---health and mortality---longer 
spacing does lead to better outcomes, although this relationship weakens with maternal 
education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

While one might expect that increasing income and female education would also lengthen
average birth spacing through access and knowledge of contraception, the empirical 
relationship between contraception use and birth spacing is ambiguous 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
To further complicate the relationship between contraception and spacing, the ability 
to successfully use ``low efficacy'' contraceptive methods, such as the rhythm method, 
is increasing in education, while there is no difference across education levels 
when it comes to modern contraceptives such as pills or injectables 
\citep{Rosenzweig1989}.
Additionally, even in the absence of modern contraceptives, families appear able to 
time or at least postpone births when they have substantial enough incentive 
\citep{Jayachandran2011,Alam2018}.

The discussion has so far assumed that children are identical---except 
for any differences that arise from spacing---but there is substantial evidence 
that the sex composition of previous children affects birth spacing, with the
absence of boys significantly decreasing the duration to next birth
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.

If short spacing lowers the survival probability of the next child, it is not, however, 
apparent that we should expect shorter spacing in the absence of boys.
There are two potential reasons why parents might still try to conceive sooner 
in the absence of boys.
One possibility is that parents do not realize that short spacing is potentially
detrimental to the next child's health, resulting in high mortality for both boys 
and girls born after short spacing across all households.

@@

@@

Women with different education levels have different hazard profiles, although the size 
and direction of the effect vary across areas and time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.
Furthermore, the use of sex selection increases with education, either because of the 
associated lower fertility or because higher income enables them to access and
afford prenatal sex determination
\citep{das_gupta97,retherford03b,Guilmoto2009a,Bongaarts2013,Portner2015b,Jayachandran2017}.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining son preference \citep{bhat03,pande07}.
}
I divide women into three groups based on education attainment: no
education, one to seven years of education, and eight and more years of education.
The models are estimated separately for each education level to reduce the potential 
problem of including other variables as proportional.

The sex composition of previous children affects both the timing of births and the use of 
sex-selective abortions 
\citep{retherford03b,Bhalotra2008,Portner2015b,Soest2018}.
I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
Finally, urban women have both lower fertility and, presumably, better access to
sex selection, both of which lead to higher use of sex-selective abortions than in 
rural areas \citep{retherford03b,Portner2015b}.
Area of residence is a dummy variable for the household living in an urban area.

@@

Appendix Tables \ref{tab:des_stat1} and \ref{tab:des_stat2} presents descriptive statistics for
the spells by education level and period.
The level of censoring increases with parity and time, as we should expect with later 
childbearing, falling fertility, and the hypothesized increases in birth intervals from 
sex-selective abortions.
